Seniority and hierarchy configuration interaction for radicals and \\ excited states

### 2023-08-23 Manuscript/manuscript.tex
ok with appendix

%%% FIG 5 %%%
\begin{figure}[h!]
\includegraphics[width=1.0\linewidth]{determinants_appendix}
\caption{
Closed-shell determinant ($s_0=0$) with $N$ electrons (left),
open-shell determinant ($s_0=1$) with $\Na$ spin-up electrons and $\Nb=\Na-1$ spin-down electrons (center),
and general determinant ($s_0=\Nas+\Nbs$), with $\Nd$ doubly-occupied orbitals, $\Nas$ singly-occupied spin-up electrons and $\Nbs$ singly-occupied spin-down electrons (right).
In the three cases, there are $K$ spatial orbitals.
The number of determinants generated from these three reference determinants, from left to right, which have excitation degree $e$ and seniority $s$
are given in Eqs.~\eqref{eq:det_1}, \eqref{eq:det_2}, and \eqref{eq:det_3}.
}
\label{fig:appendix}
\end{figure}
%%% %%% %%%

The gist of the deduction is as follows.
We start from the known number of determinants based only on the excitation degree $e$.
Then, we systematically decompose each term into specific contributions,
based on whether the excitation increases or reduces the seniority.
For the decomposition, we make use of Vandermonde's identity,
\begin{equation}
	\binom{n}{k} = \sum_{j}^{k} \binom{m}{j} \binom{n-m}{k-j}.
\end{equation}
Next, an incremental seniority can be assigned to each type of contribution.
Finally, the final answer can be obtained by summing only the contributions with the desired seniority $s$.

We start with the simpler case of a closed-shell determinant with $N$ electrons and $K$ spatial orbitals (left panel of Fig.~\ref{fig:appendix}).
From this reference determinant, the number of excited determinants generated by exciting $e$ electrons, is given by \cite{SzaboBook}
\begin{equation}
	\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p} \binom{N/2}{e-p} \binom{K-N/2}{e-p},
\label{eq:e_closed}
\end{equation}
where the sum expresses the different combinations for exciting spin-up or spin-down electrons,
the first binomial term accounts for the number of possibilities for exciting $p$ spin-up electrons from the $N/2$ orbitals,
the second term represents the number of ways of placing these $p$ electrons into $K-N/2$ orbitals,
and similarly for the latter two terms and the $e-p$ remaining spin-down electrons.

To account for the seniority of the excited determinants, one must disentangle the excitations based on how they change the seniority number,
while keeping track of the previously excited electrons.
Starting from the closed-shell determinant, $p$ spin-up electrons are excited, which increases the seniority by $2p$
(a factor of $p$ due to the unpaired spin-up electrons just excited,
and another factor of $p$ due to the unpaired spin-down electrons left behind).
The first two binomial terms in Eq.~\eqref{eq:e_closed} are left untouched since they are always accompanied by the same change in seniority.
Next, the spin-down electrons can be excited from the same orbitals from which the spin-up electrons were excited (decreasing the seniority by one)
or instead from an orbital that remained doubly-occupied (increasing the seniority by one).
The two possibilities are expressed by decomposing the third binomial term in Eq.~\eqref{eq:e_closed} as a sum over the two corresponding binomials,
\begin{equation}
	\binom{N/2}{e-p} = \sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q}.
\end{equation}
The first binomial counts the number of possibilities for exciting $q$ spin-down electrons from one of $p$ orbitals for which a spin-up electron was already excited.
By removing the unpaired spin-down electron left behind, the seniority thus decreases by a factor of $q$.
The second term accounts for the complementary excitations, where the $(e-p-q)$ spin-down electrons are chosen out of the $N/2-p$ orbitals that 
remained doubly-occupied after exciting the $p$ spin-up electrons.
Therefore, this term increases the seniority by $(e-p-q)$.
We proceed similarly for the fourth binomial term of Eq.~\eqref{eq:e_closed}, by decomposing it into the sum of two other binomials,
where $r$ spin-down electron pairs with the previously spin-up electrons (thus reducing the seniority) and $(e-p-r)$ do not (which increases the seniority).
By collecting the seniority changes from each term: $(2p)$ from the unmodified binomials, $(-q)$ and $(e-p-q)$ from the third binomial, $(-r)$ and $(e-p-r)$ from the fourth one,
the seniority of the excited determinant is given by $s = 2 (e-q-r)$.
Finally, by combining the seniority specific binomials and imposing the desired seniority via a Kronecker delta,
the number of determinants with a given excitation degree $e$ and seniority $s$ produced from a closed-shell reference ($s_0=0$) is given by
\begin{equation}
\begin{split}
	&\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p}
	\sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q}  \\
	&\sum_{r=0}^{e-p} \binom{p}{r} \binom{K-N/2-p}{e-p-r}
 	\delta_{s,2(e-q-r)}.
\label{eq:det_1}
\end{split}
\end{equation}
To obtain the final number of determinants for a given hierarchy $h$, we simply sum over the allowed combinations of $e$ and $s$ according to Eq.~\eqref{eq:h}.

Moving to the case where the number of spin-up electrons ($\Na$) and spin-down electrons ($\Nb$) differ by one ($\Na=\Nb+1$), which is illustrated in the center of Fig.~\ref{fig:appendix}.
The number of excited determinants generated exclusively by exciting $e$ electrons %$\Ndet(e)$, 
is analogous to Eq.~\eqref{eq:e_closed}, being given by
\begin{equation}
\sum_{p=0}^{e} 
\binom{\Na}{p} \binom{K-\Na}{p} \binom{\Nb}{e-p} \binom{K-\Nb}{e-p}.
\label{eq:e_open}
\end{equation}

The deduction proceeds similarly to the closed-shell case. 
The main differences are the following.
The first binomial term of Eq.~\eqref{eq:e_open} is decomposed into the sum of $t$ spin-up electrons excited from the singly-occupied orbital and $(p-t)$ that are excited from the $\Nb$ doubly-occupied ones.
The second binomial is left untouched, as before.
The third one is also decomposed as explained for the closed-shell case, with the difference that the $q$ spin-down electrons are excited from $(p-t)$ orbitals (instead of $p$) 
from which spin-up electrons were excited. 
Similarly, in the fourth binomial of Eq.~\eqref{eq:e_open}, $r$ spin-down electrons are placed into one of the $(1+p-t)$ orbitals (rather than $p$ in the closed-shell case) which contain an unpaired spin-up electron.
Collecting the individual contributions and setting the targeted seniority,
the number of determinants generated from a $s_0=1$ reference determinant, with excitation degree $e$ and seniority $s$ is given by
\begin{equation}
\begin{split}
	& \sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{1}{t} \binom{\Nb}{p-t} 
	\binom{K-\Na}{p} 
	\sum_{q=0}^{e-p} \binom{p-t}{q} \binom{\Nb-p+t}{e-p-q} \\
	& \sum_{r=0}^{e-p} \binom{1+p-t}{r} \binom{K-\Nb-1-p+t}{e-p-r}
 	\delta_{s,1+2(e-q-r-t)}.
\label{eq:det_2}
\end{split}
\end{equation}

@@

The deduction is somewhat more involved but follows along the same lines as the two previously discussed cases.
The starting point is also given by Eq.~\eqref{eq:e_open}, with each binomial being written as a sum of two or more binomials that distinguish the change in seniority.
Here, we just state how each term is decomposed, where the details can be checked by inspection of the final result presented below [see Eq.~\eqref{eq:det_3}].
The first binomial of Eq.~\eqref{eq:e_open} is decomposed into the contributions of spin-up electrons excited from the $\Nd$ doubly- and from the $\Nas$ singly-occupied orbitals.
The second binomial is decomposed into the spin-up electrons being excited to the $\Nv$ virtual orbitals and to the $\Nbs$ singly-occupied orbitals.
The third one undergoes two decompositions and is thus expressed as a double sum over three binomials.
The spin-down electrons are distinguished between the $\Nbs$ singly-occupied orbitals, a subset of $\Nd$ doubly-occupied orbitals for which spin-up electrons were excited,
and the complementary subset for which no spin-up electrons were excited.
Finally, the fourth binomial of Eq.~\eqref{eq:e_open} counts the number of possibilities to place the excited spin-down electrons.
It is first decomposed into two terms, based on whether they are promoted to the $\Nv$ virtual orbitals or to the $\Nas$ singly-occupied orbitals.
Each term is further decomposed into two terms, accounting for the subset of $\Nv$ orbitals which now contain spin-up electrons,
and the subset of $\Nas$ orbitals from which spin-up electrons were removed,
thus leading to three sums over four binomials.
The final result for the number of determinants generated from an arbitrary reference determinant, constrained to have excitation degree $e$ and seniority $s$ is given by
\begin{equation}
\begin{split}
	& \sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{\Nas}{t} \binom{\Nd}{p-t} 
	\sum_{w=0}^{p} \binom{\Nv}{w} \binom{\Nbs}{p-w} \\
	& \sum_{u=0}^{e-p} \binom{\Nbs}{u} 
	\sum_{v=0}^{e-p-u} \binom{p-t}{v} \binom{\Nd-p+t}{e-p-u-v} \\
	& \sum_{r=0}^{e-p} \sum_{m=0}^{r} \binom{w}{m} \binom{\Nv-w}{r-m} 
	\sum_{n=0}^{e-p-r} \binom{t}{n} \binom{\Nas-t}{e-p-r-n} \\
	& \delta_{s,\Nas+\Nbs+2(e+w-p-t-u-v-m-n)}.
\label{eq:det_3}
\end{split}
\end{equation}

>>>>

%%% FIG 5 %%%
\begin{figure}[h!]
\includegraphics[width=1.0\linewidth]{determinants_appendix}
\caption{
Closed-shell determinant ($s_0=0$) with $N$ electrons (left),
open-shell determinant ($s_0=1$) with $\Na$ spin-up electrons and $\Nb=\Na-1$ spin-down electrons (center),
and general determinant ($s_0=\Nas+\Nbs$), with $\Nd$ doubly-occupied orbitals, $\Nas$ singly-occupied spin-up electrons, and $\Nbs$ singly-occupied spin-down electrons (right).
In the three cases, there are $K$ spatial orbitals.
The number of determinants generated from these three reference determinants, from left to right, which have excitation degree $e$ and seniority $s$
are given in Eqs.~\eqref{eq:det_1}, \eqref{eq:det_2}, and \eqref{eq:det_3}.
}
\label{fig:appendix}
\end{figure}
%%% %%% %%%

The gist of the deduction is as follows.
We start from the known number of determinants based only on the excitation degree $e$.
Then, we systematically decompose each term into specific contributions,
based on whether the excitation increases or reduces the seniority.
For the decomposition, we make use of Vandermonde's identity:
\begin{equation}
	\binom{n}{k} = \sum_{j}^{k} \binom{m}{j} \binom{n-m}{k-j}.
\end{equation}
Next, an incremental seniority can be assigned to each type of contribution.
Finally, the final answer can be obtained by summing only the contributions with the desired seniority $s$.

We start with the simpler case of a closed-shell determinant with $N$ electrons and $K$ spatial orbitals (left panel of Fig.~\ref{fig:appendix}).
From this reference determinant, the number of excited determinants generated by exciting $e$ electrons, is given by \cite{SzaboBook}
\begin{equation}
	\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p} \binom{N/2}{e-p} \binom{K-N/2}{e-p},
\label{eq:e_closed}
\end{equation}
where the sum expresses the different combinations for exciting spin-up or spin-down electrons.
The first binomial term accounts for the number of possibilities for exciting $p$ spin-up electrons from the $N/2$ orbitals,
the second term represents the number of ways of placing these $p$ electrons into $K-N/2$ orbitals,
and similarly for the latter two terms and the $e-p$ remaining spin-down electrons.

To account for the seniority of the excited determinants, one must disentangle the excitations based on how they change the seniority number,
while keeping track of the previously excited electrons.
Starting from the closed-shell determinant, $p$ spin-up electrons are excited, which increases the seniority by $2p$
(a factor of $p$ due to the unpaired spin-up electrons just excited,
and another factor of $p$ due to the unpaired spin-down electrons left behind).
The first two binomial terms in Eq.~\eqref{eq:e_closed} are left untouched since they are always accompanied by the same change in seniority.
Next, the spin-down electrons can be excited from the same orbitals from which the spin-up electrons were excited (decreasing the seniority by one)
or instead from an orbital that remained doubly occupied (increasing the seniority by one).
The two possibilities are expressed by decomposing the third binomial term in Eq.~\eqref{eq:e_closed} as a sum over the two corresponding binomials, such as
\begin{equation}
	\binom{N/2}{e-p} = \sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q}.
\end{equation}
The first binomial counts the number of possibilities for exciting $q$ spin-down electrons from one of $p$ orbitals for which a spin-up electron was already excited.
By removing the unpaired spin-down electron left behind, the seniority thus decreases by a factor of $q$.
The second term accounts for the complementary excitations, where the $(e-p-q)$ spin-down electrons are chosen out of the $N/2-p$ orbitals that 
remained doubly-occupied after exciting the $p$ spin-up electrons.
Therefore, this term increases the seniority by $(e-p-q)$.
We proceed similarly for the fourth binomial term of Eq.~\eqref{eq:e_closed}, by decomposing it into the sum of two other binomials,
where $r$ spin-down electron pairs with the previously spin-up electrons (thus reducing the seniority) and $(e-p-r)$ do not (which increases the seniority).
By collecting the seniority changes from each term: $(2p)$ from the unmodified binomials, $(-q)$ and $(e-p-q)$ from the third binomial, $(-r)$ and $(e-p-r)$ from the fourth one,
the seniority of the excited determinant is given by $s = 2 (e-q-r)$.
Finally, by combining the seniority specific binomials and imposing the desired seniority via a Kronecker delta,
the number of determinants with a given excitation degree $e$ and seniority $s$ produced from a closed-shell reference ($s_0=0$) is given by
\begin{multline}
	\sum_{p=0}^{e} \binom{N/2}{p} \binom{K-N/2}{p}
	\sum_{q=0}^{e-p} \binom{p}{q} \binom{N/2-p}{e-p-q} 
	\\ \times
	\sum_{r=0}^{e-p} \binom{p}{r} \binom{K-N/2-p}{e-p-r}
 	\delta_{s,2(e-q-r)}.
\label{eq:det_1}
\end{multline}
To obtain the final number of determinants for a given hierarchy $h$, we simply sum over the allowed combinations of $e$ and $s$ according to Eq.~\eqref{eq:h}.

Moving to the case where the number of spin-up electrons ($\Na$) and spin-down electrons ($\Nb$) differ by one ($\Na=\Nb+1$), which is illustrated in the center panel of Fig.~\ref{fig:appendix},
the number of excited determinants generated exclusively by exciting $e$ electrons %$\Ndet(e)$, 
is analogous to Eq.~\eqref{eq:e_closed}, being given by
\begin{equation}
\sum_{p=0}^{e} 
\binom{\Na}{p} \binom{K-\Na}{p} \binom{\Nb}{e-p} \binom{K-\Nb}{e-p}.
\label{eq:e_open}
\end{equation}

The deduction proceeds similarly to the closed-shell case. 
The main differences are the following.
The first binomial term of Eq.~\eqref{eq:e_open} is decomposed into the sum of $t$ spin-up electrons excited from the singly-occupied orbital and $(p-t)$ that are excited from the $\Nb$ doubly-occupied ones.
The second binomial is left untouched, as before.
The third one is also decomposed as explained for the closed-shell case, with the difference that the $q$ spin-down electrons are excited from $(p-t)$ orbitals (instead of $p$) 
from which spin-up electrons were excited. 
Similarly, in the fourth binomial of Eq.~\eqref{eq:e_open}, $r$ spin-down electrons are placed into one of the $(1+p-t)$ orbitals (rather than $p$ in the closed-shell case) which contain an unpaired spin-up electron.
Collecting the individual contributions and setting the targeted seniority,
the number of determinants generated from a $s_0=1$ reference determinant, with excitation degree $e$ and seniority $s$ is given by
\begin{multline}
	\sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{1}{t} \binom{\Nb}{p-t} 
	\binom{K-\Na}{p} 
	\sum_{q=0}^{e-p} \binom{p-t}{q} \binom{\Nb-p+t}{e-p-q} 
	\\ \times
	\sum_{r=0}^{e-p} \binom{1+p-t}{r} \binom{K-\Nb-1-p+t}{e-p-r}
 	\delta_{s,1+2(e-q-r-t)}.
\label{eq:det_2}
\end{multline}

@@

The deduction is somewhat more involved but follows along the same lines as the two previously discussed cases.
The starting point is also given by Eq.~\eqref{eq:e_open}, with each binomial being written as a sum of two or more binomials that distinguish the change in seniority.
Here, we just state how each term is decomposed, where the details can be checked by inspection of the final result presented below [see Eq.~\eqref{eq:det_3}].
The first binomial of Eq.~\eqref{eq:e_open} is decomposed into the contributions of spin-up electrons excited from the $\Nd$ doubly- and from the $\Nas$ singly-occupied orbitals.
The second binomial is decomposed into the spin-up electrons being excited to the $\Nv$ virtual orbitals and to the $\Nbs$ singly-occupied orbitals.
The third one undergoes two decompositions and is thus expressed as a double sum over three binomials.
The spin-down electrons are distinguished between the $\Nbs$ singly-occupied orbitals, a subset of $\Nd$ doubly-occupied orbitals for which spin-up electrons were excited,
and the complementary subset for which no spin-up electrons were excited.
Finally, the fourth binomial of Eq.~\eqref{eq:e_open} counts the number of possibilities to place the excited spin-down electrons.
It is first decomposed into two terms, based on whether they are promoted to the $\Nv$ virtual orbitals or to the $\Nas$ singly-occupied orbitals.
Each term is further decomposed into two terms, accounting for the subset of $\Nv$ orbitals which now contain spin-up electrons,
and the subset of $\Nas$ orbitals from which spin-up electrons were removed,
thus leading to three sums over four binomials.
The final result for the number of determinants generated from an arbitrary reference determinant, constrained to have excitation degree $e$ and seniority $s$ is given by
\begin{widetext}
\begin{multline}
	\sum_{p=0}^{e} 
	\sum_{t=0}^{p} \binom{\Nas}{t} \binom{\Nd}{p-t} 
	\sum_{w=0}^{p} \binom{\Nv}{w} \binom{\Nbs}{p-w} 
	\sum_{u=0}^{e-p} \binom{\Nbs}{u} 
	\sum_{v=0}^{e-p-u} \binom{p-t}{v} \binom{\Nd-p+t}{e-p-u-v} 
	\sum_{r=0}^{e-p} \sum_{m=0}^{r} \binom{w}{m} \binom{\Nv-w}{r-m} 
	\\ \times
	\sum_{n=0}^{e-p-r} \binom{t}{n} \binom{\Nas-t}{e-p-r-n} 
	\delta_{s,\Nas+\Nbs+2(e+w-p-t-u-v-m-n)}.
\label{eq:det_3}
\end{multline}
\end{widetext}
